---
title: "Semiparametric Regression - Assignment 5"
author: "Szymon Armata <br/>| 341593 <br/>| Data Science"
date: "15 November 2022"
output:
  pdf_document:
    toc: yes
    number_sections: yes
    latex_engine: xelatex
    extra_dependencies: subfig
header-includes: \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
knitr::opts_chunk$set(fig.pos = 'H')
```
\newpage

\section{Plot of the data}
```{r, echo=TRUE}
library(nlme)
library(HRW)
data(WarsawApts)

data(WarsawApts)
x <- WarsawApts$construction.date
y <- WarsawApts$areaPerMzloty
```
```{r, echo=TRUE}
plot(x,y,
     xlab = "construction date",
     ylab = "area per million zloty", col="dodgerblue",
     main = "Warsaw apartments: area vs. construction date")
```

\newpage
\section{Task 3.}
\subsection{a}
```{r, echo=TRUE}
area.perMz <- WarsawApts$areaPerMzloty
const.date <- WarsawApts$construction.date
```
\subsection{b}
```{r, echo=TRUE}
numObs <- length(const.date)
X <- cbind(rep(1,numObs), const.date)
```
\subsection{c}
```{r, echo=TRUE}
numIntKnots <- 35
intKnots <- quantile(unique(const.date),
seq(0, 1,
length = numIntKnots + 2))[-c(1,numIntKnots + 2)]
Z <- outer(const.date, intKnots, "-")
Z <- Z * (Z > 0)
```
\subsection{d}
```{r, echo=TRUE}
dummyId <- factor(rep(1, numObs))
Z.sm <- list(dummyId = pdIdent(~ -1 + Z))
fit <- lme(area.perMz ~ -1 + X, random = Z.sm)
```

\newpage
\section{Task 4.}
\subsection{a}
```{r, echo=TRUE}
ng <- 1001
range.date <- range(const.date)
dategrid <- seq(range.date[1], range.date[2], length = ng)
Xg <- cbind(rep(1, ng), dategrid)
Zg <- outer(dategrid, intKnots, "-")
Zg <- Zg * (Zg > 0)
```
\subsection{b}
```{r, echo=TRUE}
betaHat <- as.vector(fit$coef$fixed)
uHat <- as.vector(fit$coef$random[[1]])
```
\subsection{c}
```{r, echo=TRUE}
fhat <- Xg %*% betaHat + Zg %*% uHat
```
\subsection{d}
```{r, echo=TRUE}
plot(const.date, area.perMz,
xlab = "Construction date (years)",
ylab = "Area per million zloty (m2)",
main = "Warsaw apartments: area vs. construction date")
lines(dategrid, fhat, lwd = 2, col = "red")
```

Even though there are times when there is a shortage of data (for example, the years 1940-1945 or 1980–1995) and the line appears to be overfitted, it can be seen that the fitting line represents the trend in apartment prices.
